### 2021-11-17 plos_latex_template.tex
Rocío: edits to the intro and fig. 1

\newcommand{\fede}[1]{\textcolor{ForestGreen}{Fede: #1}}
\newcommand{\rocio}[1]{\textcolor{MidnightBlue}{Rocio: #1}}
\newcommand{\doro}[1]{\textcolor{RedViolet}{Dorothea: #1}}
\newcommand{\jani}[1]{\textcolor{YellowOrange}{Janina: #1}}
\newcommand{\la}[1]{\textcolor{RawSienna}{Laura: #1}}
\newcommand{\as}[1]{\textcolor{Violet}{Andrea: #1}}

@@

% Rocío: Main message of the paragraph: Why we are writing this piece
Conferences, from the Latin \textit{conferre}--`to bring together', are spaces to meet and reconnect with members from a specific community, learn about advances in the field, and share our recent contributions.\fede{between this and the next sentence, there is a strong break. need some ligation, something like "Despite the many efforts in this direction, ..."}
However, conferences are likely to reproduce the systematic discrimination occurring in other spaces in our fields. 
Lack of representation and unwelcoming--or overtly aggressive--environments exclude people from opportunities for learning, sharing, and producing knowledge, and becoming active community members \cite{hendersonThoughtfulGatheringsGendering2020}.
Exclusionary conference experiences can divert career paths, affect lives, and drive people out of academia \cite{biggsAcademicConferenceChilly2018, hendersonThoughtfulGatheringsGendering2020}. 
This message is not new. 
Barriers such as unattainable registration costs, sexism, prejudice against disable people (ableism), among others, have already been exposed and discussed in the literature \cite{arendDisparityConferenceRegistration2019, biggsAcademicConferenceChilly2018, depickerRethinkingInclusionDisability2020a, irishIncreasingParticipationUsing2020}, 
and some proposals for more inclusive conferences have been put into practice \cite{gichoraTenSimpleRules2010a, levitisCenteringInclusivityDesign2021, atkinsonJournalMedicine20202021, foramittiVirtuesVirtualConferences2021, ninerBetterWhomLeveling2021, rabyMovingAcademicConferences2021, noauthor_discover2021},
with a primary focus on the online format to open doors for inclusion.\fede{do we need maybe a short definition of the concepts of diversity-equity-inclusion for us/refer to something?} %LA: https://www.cscce.org/resources/organizing-community-events/

% Rocío: Main message of the paragraph: Roadmap of the paper
This article focuses on practices for hosting inclusive conferences.
The rules written here are directed to people who are part of a stable meetings committee that oversees the site/location selection process, or that coordinates with the local organizers of conferences.
The rules can also be helpful to local/virtual organizers who intend to put inclusion at the core of the planning phase.
These tips stem from the authors' experience of organizing useR! 2021, a virtual and global statistical computing conference for users and developers of the R programming language \cite{r_core_team_2021}. 
We embraced the challenge of organizing a high-quality virtual conference in the context of the COVID-19 pandemic and making it a kind, inclusive, and accessible experience for as many persons as possible. 
Here, we share the lessons learned within the time organizing useR! 2021, summarized as ten simple rules towards hosting an inclusive conference.
The rules are organized in three groups (Figure \ref{fig:diagram}). %original source: https://docs.google.com/drawings/d/1iS1pLc9OldLMe_jhNT6OGIbT9s5KvlHpUoDd19vHkCY/edit?usp=sharing
%sara: still need to better format figure size and get the alt text right :P
Group 1 includes rules 1, 2, and 10, which refer to pillars of an inclusive conference: embracing diversity in all its dimensions, creating a safe and welcoming environment for everyone, and making the conference part of a long-term process for inclusion.
The second group includes rules 3 and 4 and focuses on the people who participate in the conference. 
Rule 3 refers to the importance of working with an inclusive and diverse organizing team, and Rule 4 concerns the necessity of counteracting implicit and systemic bias from spotlight roles like keynote speakers, other presenters, program committee members, or other session chairs. 
The third group includes rules 5 to 9. These rules are about components of the conference that should be carefully planned for: an online component, accessibility for people with disabilities, language inclusiveness, a welcoming communication strategy, and financial resources to support inclusion. 
These rules apply equally to conferences where the expected audience is academic, from the industry, or a mixed group. While these rules are inspired in a global virtual experience they also apply at the regional or local level and can inform in-person and hybrid events.
\fede{maybe something like: "Even if this set of rules is derived from our experience with a global scale event, we believe these can be easily transferred to events with regional or local scopes"}

\begin{figure}[!h]
\centering
%sara: without success i'm trying to use pdfcomment package and pdftooltip to make the alt-text. any help?
\pdftooltip{
\includegraphics[width=\textwidth]{figs/10_rules_diversity.png}}{
%sara: suggestion for alt text:
Diagram of how the 10 rules are organized into three groups: grounding rules (rules 1, 2, and 10), people-related rules (rules 3 and 4), and design rules (rules 5 to 9). The diagram has five rows. Each line contains rectangular boxes with the rule number and a short title for each rule. Each rule box is colored based on their group (grounding rules are grey, people-related rules are green and design rules are pink). First row is for the grounding rules 1. Multidimensional diversity and 2. Safe and inclusive environment in grey boxes. Second row is for people-related rules 3. Organizing team and 4. Unbias in green boxes. Third row is for the design rules 5. Online component, 6. Accessibility for disabilities, 7. Language inclusivity in pink boxes. The fourth row is for the design rules 8. Communication and 9. Financial support and budget in pink boxes. The last row is for the grounding rule 10. Diversity and inclusion as a process in grey boxes.
}
\caption{Diagram of the rules organized in three groups: grounding rules (rules 1, 2, and 10), people-related rules (rules 3 and 4), and design rules (rules 5 to 9).}
\label{fig:diagram}
\end{figure}

@@

% Rocío: Main idea: what is diversity and why does it matter? (inequalities)
Diversity encompasses multiple dimensions: age, physical ability, career stage, gender, gender identity, geographic origin, language, neurodiversity, race, religion, sexual orientation, and socioeconomic background, to name a few.
Human diversity should be celebrated and respected in every way. Nonetheless, we live in a world with implicit hierarchies along these axes. Some statuses (e.g., cisgender, white, male, from the United States or Western Europe) hold the privilege of being the defaults around which all systems--including conferences--are consciously and unconsciously built. \as{People outside these groups suffer different forms of oppression, like sexism, racism, ableism, homophobia, and transphobia. Furthermore, no diversity dimension operates in isolation from the others, they intersect in complex ways to determine people's experience in the world--a concept know as \textit{intersectionality}--so that anyone can suffer from some forms of oppression and be privileged along some other dimension, and some groups are particularly vulnerable and systematically left behind.}
While no isolated initiative can change reality by itself, building a more diverse and inclusive conference starts by recognizing that these inequalities have systematically excluded whole groups of people from academia and scientific and professional circles \cite{timperleyHeMoanaPukepuke2020}.

% Rocío: more effort towards including the more excluded
Recognizing our privileges--unearned advantages given by society to some people but not all--particularly in our field and in our scientific or professional community, will help identify which subgroups have been the most excluded or discriminated against. 
These are the groups we need to make more effort to include.
Investing more effort towards the most excluded groups does not mean neglecting the others, but it will guide the vision of diversity for your conference--and your strategies to achieve it.
This may translate into different actions depending on your field, region, or community, e.g. a gender distribution of your speakers that is representative of the general population, presence of racialized people --especially Black people-- among the head organizers, speakers, and attendees, LGBTQIA+ friendly-spaces, or community participation from key geographic regions.

@@

A genuinely inclusive conference can only be organized by an inclusive and diverse organizing team, in charge of the conference decision-making.
Organizers should build a team with people from different regions, genders, races and ethnicities, socioeconomic statuses, and other aspects of diversity.
Having diverse people in decision-making positions will affect positively all the other aspects of your conference because all the processes will benefit from their input, expertise, and distinct perspectives \cite{hongGroupsDiverseProblem2004}. 
In addition, a diverse team plays an important role at creating a welcoming space because representation--seeing people with similar life experiences occupy public spaces, positions of power, and breaking negative stereotypes--is one of the best ways to create a sense of belonging for everyone participating in the conference (see \textbf{Rule \ref{rule_inclusion}}). If you already started assembling an organizing team, check for gaps in its composition. 
The regional and local communities, groups, or associations in your field are good sources to tap into.

% Rocío: Idea of this paragraph: non-transferability and do not restrict people
People with disabilities often say: `Nothing about us without us'; the same holds for other dimensions of diversity. This means that the actual life experiences, expertise, and insights from people in marginalized groups cannot be replaced by good intentions from people outside these groups \cite{costanzachockDesign2020}.
People who have experienced exclusion have the social and technical expertise to fight against it and they will be essential in any team.
However, they must have the freedom to choose in which areas of the conference they want to work and not be restricted to diversity and inclusion aspects.

% Rocío: Idea of this paragraph: tips to make a diverse and inclusive team work + paying them
When working with people from marginalized groups do not assume that self-nomination and voting will work as mechanisms to choose leadership positions. Instead, nominate directly and offer leading positions that would normally be occupied by people from privileged groups.
Build an environment in which every person can express their position and give priority to people from systematically excluded groups.
Many people lack the institutional support to put time and effort into the organization tasks and do not have the luxury to commit to the organization for free; consider paying them as an item in your budget.  
% Saranjeet: At times regional institutes and universities do not support nor recognize the time put in organization tasks. As such one needs to carry out their day work for the institute while working extra hours for organizing. It might be challenging to find a way to recognize organizers that works internationally. However, there should certainly be some mechanism put into place which recognizes the time of the organizers.
% Rocío: Add other types of recognition besides money (website, public recognition on social media, during the conference, etc.)
In addition, tasks such as receiving and responding to Code of Conduct reports can be emotionally intense work and should be additionally rewarded.

@@

\section{Consciously counteract bias in the conference program}
\label{rule_unbias}

%idea of paragraph 1: our lists are biased

@@

% our committees are biased and unconscious bias exists
Having a chance to present at a conference is an opportunity to gain--or keep--visibility in the community or field. 
Even diverse selection committees can be subjected to unconscious bias, by interpreting information such as the location of the institution or the names of the authors and attaching value to their work according to negative or positive stereotypes. 
Selection committees need to be reminded of such biases, and try to counteract them when evaluating submissions and presentations, which should have clear evaluation criteria.
\cite{swartzScienceValueDiversity2019, wongBuildDiversityScience2020}

% Rocío: main idea: unbias roles and give spotlight to the roles--and people who have contributed in roles--that are not stereotypically categorized as success
Activities such as scientific publication and development of software are often regarded as higher value than care-taking roles like community building or teaching. 
These roles may be equally if not more challenging and are usually assumed by women, people of color, people with disabilities, and other minoritized groups \cite{cheng2020x+, burfordHomelinessMeantHaving2020}. Research about these topics is equally undervalued compared to other areas of knowledge production. 
Counteracting bias also means broadening the range of topics of the conference and giving spotlight/visibility to usually undervalued research and the people behind it.
%Defy the stereotypical criteria for merit and how it is traditionally measured by acknowledging these community practices and the people behind them.
You can also reframe the awards ceremony to acknowledge those who contributed to community building, presented teaching materials, and assumed other care-taking roles.

@@

% Rocío: main idea: Networking may seem the weakest part of online conference, but it doesn't have to be. 
Networking and socializing have been mentioned as challenging aspects of online conferences, mostly because we are used to interactions at in-person social settings such as coffee or lunch breaks \cite{salibaGettingGripsOnline2020, roosOnlineConferencesNew2020}. 
However, in-person settings might not be comfortable or appealing to everyone for socializing and some of these spaces may be exclusionary (e.g. galas or dinner nights at expensive venues). 
On the other hand, there is evidence that virtual communication before face-to-face interaction can make people from minoritized groups feel more included, thus participate more (e.g. \cite{trianaDoesOrderFacetoFace2012, blackEngenderingBelongingThoughtful2020}). 
Organizers of online conferences should invest time in creating opportunities to meet and bond virtually, respecting people's limits, preferences, and remembering that `the usual' does not necessarily work for everyone, and that no single networking activity will ever serve the whole community. 
It is worth trying new and varied activities centered around different groups of people outside the mainstream of your conference.
Some ideas in this line are: offering the option of written chat instead of voice or video conversations, opening events with teamwork like trivia, offer some events that can be enjoyed passively like movies, yoga sessions, or art displays, where attendants can choose to just sit and enjoy without talking, or have a chat channel to comment on their experiences during the session.\fede{another activity: virtual city tours, I had it as a corollary activity for a summer school} \la{Some more specifics on how to achieve these suggestions in this paper: https://osf.io/k3bfn/}
% we need to check if something here (newbies!) is not missing in the above paragraphs
% As we mentioned previously, conference attendees should have ample opportunities to network with each other. However, having a diverse offer of networking opportunities that can appeal to people with different backgrounds, accessibility needs, and preferences can be challenging, especially if you have the mindset of organizing each activity for the complete pool of attendants. If you think of smaller activities that can reach specific groups, preferably co-led by community leaders, these sessions can be more productive successful, and inclusive than trying to organize one single activity that can please the whole community (\textbf{Rule \ref{rule_unbias}}). Some examples are the newbies sessions for first-timers, mixers lead by specific subgroups or communities, or leisure activities that reunite subgroups with the same interests: arts, exercise, sports, movies, etc.

% Rocío: main idea of the paragraph: In the context of reopening, hybrid conferences may be a thing. They also need to be inclusive.
Within the context of the COVID-19 pandemic, many conferences embraced the online format; but at the time of writing, some are reverting to in-person, which risks going back to the barriers mentioned above \cite{jooKeepOnlineOption2021}.
Alternatively, a conference could have a hybrid format with an in-person and an online components that articulate well and maximize the experience of everyone attending your conference. This dual format could allow a group of people to interact face-to-face while providing many others the opportunity to participate remotely. 
The challenge and requirement for this kind of setting would be to make the online component as relevant as the in-person component and not just a consolation prize to the less privileged in the community \cite{jooKeepOnlineOption2021,ninerBetterWhomLeveling2021}.

In fact, experts in theorizing about and building communities of practice [Wenger, 1999] who have been holding hybrid events since 2011 [Wenger-Tayner, 2021] consider that inclusive hybrid events are the future. These authors emphasize the need that all participants of hybrid events must go through the same event quality to the point that in-person and online participants pay the same registration amount to participate in high-quality hybrid conferences. These authors recommend having a buddy system, where every person attending online has an in-person attendant that makes sure the online person can follow the event and participate in the event as if they were in the room [Wenger-Trayner, 2020]. Wenger-Trayner (2020, 2021) also recommends having a chat function for back-channel conversations between all attendants, a virtual space for all online attendants to talk to each other, and allocating time to small group conversations, often mixing in-person and online groups. The physical venue will have to include multiple ways to connect with online folks such as tablets on the meeting venue tables and cameras that allow online attendants to follow who is speaking in the in-person space.

>>>>

\newcommand{\fede}[1]{\textcolor{ForestGreen}{Fede: #1}}
\newcommand{\rocio}[1]{\textcolor{MidnightBlue}{Rocio: #1}}
\newcommand{\doro}[1]{\textcolor{RedViolet}{Dorothea: #1}}
\newcommand{\jani}[1]{\textcolor{YellowOrange}{Janina: #1}}
\newcommand{\la}[1]{\textcolor{RawSienna}{Laura: #1}}
\newcommand{\as}[1]{\textcolor{Violet}{Andrea: #1}}
\newcommand{\rita}[1]{\textcolor{CarnationPink}{Rita: #1}}

@@

% Rocío: Main message of the paragraph: Conferences are great, but for some
Conferences, from the Latin \textit{conferre}--`to bring together', are spaces to meet and reconnect with members from a specific community, learn about advances in the field, and share our recent contributions.
The larger the conference, the larger the opportunities to network and learn from the cohort.
However, opportunities for participating in conferences are not equal for all. 
Just like in academia and tech in general, conferences can become exclusive spaces for privileged groups (i.e. with unearned advantages given by society to some but not all, e.g. white, male, from a rich country, English-native speaker, with no physical disabilities) (\cite{arendDisparityConferenceRegistration2019, biggsAcademicConferenceChilly2018, depickerRethinkingInclusionDisability2020a, irishIncreasingParticipationUsing2020}).
% However, conferences are likely to reproduce the systematic discrimination occurring in other spaces in our fields.

This article proposes 10 rules to host inclusive conferences.
They are organized in three sections: the pillars (embracing diversity, creating a welcoming environment, and having a diverse organizing team), design rules (counteracting bias in the conference program),  (having a strong online component, achieving accessibility for people with disabilities, embracing language diversity, having inclusive communication strategies, and allocating financial resources to support inclusive practices), and a process rule: make the conference part of a long-term process for inclusion. 
These practices stem from the experience of organizing useR! 2021, a virtual and global statistical computing conference for users and developers of the R programming language \cite{r_core_team_2021}. 
We embraced the challenge of organizing a high-quality virtual conference in the context of the COVID-19 pandemic and making it a kind, inclusive, and accessible experience for as many persons as possible. 
The list of authors is composed of the global coordinators of the conference, members of the diversity and inclusion team, and people who contributed to other areas of the conference. 
Therefore, the rules are a result of our lessons learned from diverse perspectives and expertise (i.e. our different backgrounds, profiles, and roles in the conference). 
Even if this set of rules is derived from our experience with a global scale event, we believe these can be easily transferred to events with regional or local scopes.
They are directed to people who are part of a stable meetings committee that oversees the site/location selection process, or that coordinates with the local organizers of conferences.
The rules can also be helpful to local/virtual organizers who intend to put inclusion at the core of the planning phase.

% This message is not new. 
% Barriers such as unattainable registration costs, sexism, prejudice against disable people (ableism), among others, have already been exposed and discussed in the literature \cite{arendDisparityConferenceRegistration2019, biggsAcademicConferenceChilly2018, depickerRethinkingInclusionDisability2020a, irishIncreasingParticipationUsing2020}, 
% %timperleyHeMoanaPukepuke2020, gewinWhatScientistsShould2019, brownAbleismAcademiaWhere2018, marks2021meeting
% and some proposals for more inclusive conferences have been put into practice \cite{gichoraTenSimpleRules2010a, levitisCenteringInclusivityDesign2021, atkinsonJournalMedicine20202021, foramittiVirtuesVirtualConferences2021, ninerBetterWhomLeveling2021, rabyMovingAcademicConferences2021, noauthor_discover2021},
% with a primary focus on the online format to open doors for inclusion.\fede{do we need maybe a short definition of the concepts of diversity-equity-inclusion for us/refer to something?} %LA: https://www.cscce.org/resources/organizing-community-events/
% %By inclusion, we mean creating conferences organized by and dedicated to a group of attendees that is representative of the whole population. This means creating a welcoming, respectful, and safe conference that tries to counteract systemic oppression.

% % Rocío: Main message of the paragraph: Roadmap of the paper
% This article focuses on practices for hosting inclusive conferences.
% The rules written here are directed to people who are part of a stable meetings committee that oversees the site/location selection process, or that coordinates with the local organizers of conferences.
% The rules can also be helpful to local/virtual organizers who intend to put inclusion at the core of the planning phase.
% These tips \rita{tips looks to me too informal for an article, maybe used the word `suggestions` or `advice`?} stem from the authors' experience of organizing useR! 2021, a virtual and global statistical computing conference for users and developers of the R programming language \cite{r_core_team_2021}. 
% We embraced the challenge of organizing a high-quality virtual conference in the context of the COVID-19 pandemic and making it a kind, inclusive, and accessible experience for as many persons as possible. 
% Here, we share the lessons learned within the time organizing useR! 2021, summarized as ten simple rules towards hosting an inclusive conference.
% The rules are organized in three groups (Figure \ref{fig:diagram}). %original source: https://docs.google.com/drawings/d/1iS1pLc9OldLMe_jhNT6OGIbT9s5KvlHpUoDd19vHkCY/edit?usp=sharing
% %sara: still need to better format figure size and get the alt text right :P

% Group 1 includes rules 1, 2, and 10, which refer to pillars of an inclusive conference: embracing diversity in all its dimensions, creating a safe and welcoming environment for everyone, and making the conference part of a long-term process for inclusion.
% The second group includes rules 3 and 4 and focuses on the people who participate in the conference. 
% Rule 3 refers to the importance of working with an inclusive and diverse organizing team, and Rule 4 concerns the necessity of counteracting implicit and systemic bias from spotlight roles like keynote speakers, other presenters, program committee members, or other session chairs. 
% The third group includes rules 5 to 9. These rules are about components of the conference that should be carefully planned for: an online component, accessibility for people with disabilities, language inclusiveness, a welcoming communication strategy, and financial resources to support inclusion. 
% These rules apply equally to conferences where the expected audience is academic, from the industry, or a mixed group. While these rules are inspired in a global virtual experience they also apply at the regional or local level and can inform in-person and hybrid events.
% \fede{maybe something like: "Even if this set of rules is derived from our experience with a global scale event, we believe these can be easily transferred to events with regional or local scopes"}

\begin{figure}[!h]
\centering
%sara: without success i'm trying to use pdfcomment package and pdftooltip to make the alt-text. any help?
%%JPNG am thinking of the metaphor of gears in a machine as a proxy for the process.  Think also about the inclusiveness of the different rules and how some facilitate others. The most encompassing rule is the number 10. This rule recognises the community learning process of DEI instead of being solely a practical and handy way of doing things. This should be the more comprehensive and inclusive level (gear) that include all others. Rules 1 and 2 are crucial because they represent two of the most important concepts we need to grasp: diversity and the requirements to express it freely. Rules 3 to 9 are the necessary practical pre-conditions that make the community learning process about DEI possible in the framework of a conference. I think that rules 1 and 2 can be represented by bigger toothed wheels than rules 3 to 9, but all need to fit and function together.The teeth of the toothed wheels would be representations of human forms representing the diverse teams involved.
\pdftooltip{
\includegraphics[width=\textwidth]{figs/rules.png}}{
%sara: suggestion for alt text:
%Diagram of how the 10 rules are organized into three groups: grounding rules (rules 1, 2, and 10), people-related rules (rules 3 and 4), and design rules (rules 5 to 9). The diagram has five rows. Each line contains rectangular boxes with the rule number and a short title for each rule. Each rule box is colored based on their group (grounding rules are grey, people-related rules are green and design rules are pink). First row is for the grounding rules 1. Multidimensional diversity and 2. Safe and inclusive environment in grey boxes. Second row is for people-related rules 3. Organizing team and 4. Unbias in green boxes. Third row is for the design rules 5. Online component, 6. Accessibility for disabilities, 7. Language inclusivity in pink boxes. The fourth row is for the design rules 8. Communication and 9. Financial support and budget in pink boxes. The last row is for the grounding rule 10. Diversity and inclusion as a process in grey boxes.
}
\caption{Diagram of the rules organized in three groups: the pillars (rules 1, 2, and 3), design rules (rules 4 to 9), and the process (rule 10).}
\label{fig:diagram}
\end{figure}

@@

% Rocío: Main idea: what is diversity and why does it matter? (inequalities)
Diversity encompasses multiple dimensions: age, physical ability, career stage, gender, gender identity, geographic origin, language, neurodiversity, race, religion, sexual orientation, and socioeconomic background, to name a few.
Human diversity should be celebrated and respected in every way. Nonetheless, we live in a world with implicit hierarchies along these axes. Some statuses (e.g., cisgender, white, male, from the North America or Western Europe) hold the privilege of being the defaults around which all systems--including conferences--are consciously and unconsciously built. \as{People outside the dominant groups suffer different forms of oppression, like sexism, racism, ableism, homophobia, and transphobia. Furthermore, no diversity dimension operates in isolation from the others, they intersect in complex ways to determine people's experience in the world--a concept know as \textit{intersectionality}--so that anyone can suffer from some forms of oppression and be privileged along some other dimension, and some groups are particularly vulnerable and systematically left behind.}
While no isolated initiative can change reality by itself, building a more diverse and inclusive conference starts by recognizing that these inequalities have systematically excluded whole groups of people from academia and scientific and professional circles \cite{timperleyHeMoanaPukepuke2020}.

% Rocío: more effort towards including the more excluded
Recognizing our privileges--unearned advantages given by society to some people but not all--particularly in our field and in our scientific or professional community, will help identify which subgroups have been the most excluded or discriminated against. 
These are the groups we need to make more effort to include.
Investing more effort towards the most excluded groups does not mean neglecting the others, but it will guide the vision of diversity for your conference--and your strategies to achieve it.
This may translate into different actions depending on your field, region, or community, e.g. a gender distribution of your speakers that is representative of the general population, presence of racialized people --especially Black people-- among the head organizers, speakers, and attendees, LGBTQIA+ friendly-spaces, or community participation from key geographic regions. For academic conferences consider also to select for a talk master and PhD student and not to give voice only to senior scientist or principal investigator.

@@

A genuinely inclusive conference can only be coordinated  by an inclusive and diverse organizing team, in charge of the conference decision-making.
Organizers should build a team with people from different regions, genders, races and ethnicities, socioeconomic statuses, and other aspects of diversity.
Having diverse people in decision-making positions will affect positively all the other aspects of your conference because all the processes will benefit from their input, expertise, and distinct perspectives \cite{hongGroupsDiverseProblem2004}. 
In addition, a diverse team plays an important role at creating a welcoming space because representation--seeing people with similar life experiences occupy public spaces, positions of power, and breaking negative stereotypes--is one of the best ways to create a sense of belonging for everyone participating in the conference (see \textbf{Rule \ref{rule_inclusion}}). If you already started assembling an organizing team, check for gaps in its composition. 
The regional and local communities, groups, or associations in your field are good sources to tap into.

% Rocío: Idea of this paragraph: non-transferability and do not restrict people
People with disabilities often say: `Nothing about us without us'; the same holds for other dimensions of diversity. This means that the actual life experiences, expertise, and insights from people in marginalized groups cannot be replaced by good intentions from people outside these groups \cite{costanzachockDesign2020}.
People who have experienced exclusion have the social and technical expertise to fight against it and they will be essential in any team.
However, they must have the freedom to choose in which areas of the conference they want to work and not be restricted to diversity and inclusion aspects. 
%ast: i dont know if adding something about the other side of the spectrum would work, something in the lines of "On the other hand, the non-transferability of abilities and experiences does not mean that people from privileged groups cannot work together to fight all forms of oppression. on the contrary," using power, co-liberation logic.
% Rocío: Idea of this paragraph: tips to make a diverse and inclusive team work + paying them
When working with people from marginalized groups do not assume that self-nomination and voting will work as mechanisms to choose leadership positions. Instead, nominate directly and offer leading positions that would normally be occupied by people from privileged groups.
Build an environment in which every person can express their position and give priority to people from systematically excluded groups.
Many people lack the institutional support to put time and effort into the organization tasks and do not have the luxury to commit to the organization for free; consider paying them as an item in your budget.  
% Saranjeet: At times regional institutes and universities do not support nor recognize the time put in organization tasks. As such one needs to carry out their day work for the institute while working extra hours for organizing. It might be challenging to find a way to recognize organizers that works internationally. However, there should certainly be some mechanism put into place which recognizes the time of the organizers.
% Rocío: Add other types of recognition besides money (website, public recognition on social media, during the conference, etc.)
In addition, tasks such as receiving and responding to Code of Conduct reports can be emotionally intense work and should be additionally rewarded.

@@

\section{Consciously counteract bias in the conference program}
\label{rule_unbias}
%la: LA: repeating parts of my comment in Figure 1:maybe "mitigate bias" - unbiasedness is a reasonable goal once systemic biases are gone, in the meantime, introducing biases favoring marginalized folks is required to reach equity. Thus, while unbiasedness is the ultimate goal, the current rules should point to mitigate biases and that may require to bias in favor of marginalized groups. Equity and Justice call for biases in favor of marginalized folks. Unbiasing aims for equality that, in the current status quo, will keep favoring those who are already in the mainstream and will keep increasing marginalization for those currently in the margins. This is a great reference regarding the framework "mainstream/margins" https://www.trainingforchange.org/wp-content/uploads/2017/11/Mainstream-Margin-in-Groups.pdf See the cartoon about equality, equity, and justice here: https://onlinepublichealth.gwu.edu/resources/equity-vs-equality/ and here: https://www.reddit.com/r/coolguides/comments/o3ivsg/equality_equity_and_justice_explained_better/
%idea of paragraph 1: our lists are biased

@@

% our committees are biased and unconscious bias exists
Having a chance to present at a conference is an opportunity to gain--or keep--visibility in the community or field. %ast is something missing in this transition?

Even diverse selection committees can be subjected to unconscious bias, by interpreting information such as the location of the institution or the names of the authors and attaching value to their work according to negative or positive stereotypes. 
Selection committees need to be reminded of such biases, and try to counteract them when evaluating submissions and presentations, which should have clear evaluation criteria.
\cite{swartzScienceValueDiversity2019, wongBuildDiversityScience2020}

% Rocío: main idea: unbias roles and give spotlight to the roles--and people who have contributed in roles--that are not stereotypically categorized as success
Activities such as scientific publication and development of software are often regarded as higher value than care-taking roles like community building or teaching. 
These roles may be equally if not more challenging and are usually assumed by women, people of color, people with disabilities, and other minoritized groups \cite{cheng2020x+, burfordHomelinessMeantHaving2020}. Research about these topics is equally undervalued compared to other areas of knowledge production. 
Counteracting bias also means broadening the range of topics of the conference and giving spotlight/visibility to usually undervalued research and the people behind it.
%Defy the stereotypical criteria for merit and how it is traditionally measured by acknowledging these community practices and the people behind them.
You can also reframe the awards ceremony \as{to acknowledge work beyond the traditional definition of merit/quality and include} those who contributed to community building, presented teaching materials, and assumed other care-taking roles.

@@

%ast add something about online giving flexibility in formats and how that depends on to the context of the conference
% Rocío: main idea: Networking may seem the weakest part of online conference, but it doesn't have to be. 
Networking and socializing have been mentioned as challenging aspects of online conferences, mostly because we are used to interactions at in-person social settings such as coffee or lunch breaks \cite{salibaGettingGripsOnline2020, roosOnlineConferencesNew2020}. 
However, in-person settings might not be comfortable or appealing to everyone for socializing and some of these spaces may be exclusionary (e.g. galas or dinner nights at expensive venues). 
On the other hand, there is evidence that virtual communication before face-to-face interaction can make people from minoritized groups feel more included, thus participate more (e.g. \cite{trianaDoesOrderFacetoFace2012, blackEngenderingBelongingThoughtful2020}). 
%JPNG: I think in-person networking during a conference has a disadvantage in contrast to online and virtual interactions. In-person networking suffers from being contingent; they are meaningful only to the people that happen to be in the same place and at the precise time. If you weren't there during the coffee break, you lost the networking opportunity. On the contrary, written media like chats are recorded and stored, allowing you to join the conversation at every time while catching up on what has been said. This creates the opportunity for long term interactions during the conference, which is good for everyone and not only for people with some disadvantages (language) or different personality traits (introversion, shyness, etc.). I saw this kind of interaction not only during useR!2021 but also in other virtual conferences I had the opportunity to assist with. If you think this idea is relevant, we could add something like: "On the other hand, there is evidence that virtual communication before face-to-face interaction can make people from minoritized groups feel more included, thus participate more (e.g. [29, 30]). Written media like chats create opportunities for long term interactions and networking by allowing more people to join the conversation even when it has started some time ago."
Organizers of online conferences should invest time in creating opportunities to meet and bond virtually, respecting people's limits, preferences, and remembering that `the usual' does not necessarily work for everyone, and that no single networking activity will ever serve the whole community. 
It is worth trying new and varied activities centered around different groups of people outside the mainstream of your conference.
%la: see https://www.trainingforchange.org/wp-content/uploads/2017/11/Mainstream-Margin-in-Groups.pdf
Some ideas in this line are: offering the option of written chat in addition to voice or video conversations, opening events with teamwork like trivia, offer some events that can be enjoyed passively like movies, yoga sessions, or art displays, where attendants can choose to just sit and enjoy without talking, or have a chat channel to comment on their experiences during the session.\fede{another activity: virtual city tours, I had it as a corollary activity for a summer school} \la{Some more specifics on how to achieve these suggestions in this paper: https://osf.io/k3bfn/}
% we need to check if something here (newbies!) is not missing in the above paragraphs
% As we mentioned previously, conference attendees should have ample opportunities to network with each other. However, having a diverse offer of networking opportunities that can appeal to people with different backgrounds, accessibility needs, and preferences can be challenging, especially if you have the mindset of organizing each activity for the complete pool of attendants. If you think of smaller activities that can reach specific groups, preferably co-led by community leaders, these sessions can be more productive successful, and inclusive than trying to organize one single activity that can please the whole community (\textbf{Rule \ref{rule_unbias}}). Some examples are the newbies sessions for first-timers, mixers lead by specific subgroups or communities, or leisure activities that reunite subgroups with the same interests: arts, exercise, sports, movies, etc.

% Rocío: main idea of the paragraph: In the context of reopening, hybrid conferences may be a thing. They also need to be inclusive.
Within the context of the COVID-19 pandemic, many conferences embraced the online format; but at the time of writing, some are reverting to in-person, which risks going back to the barriers mentioned above \cite{jooKeepOnlineOption2021}.
Alternatively, a conference could have a hybrid format with an in-person and an online components that articulate well and maximize the experience of everyone attending your conference. This dual format could allow a group of people to interact face-to-face while providing many others the opportunity to participate remotely. 
The challenge and requirement for this kind of setting would be to make the online component as relevant as the in-person component and not just a consolation prize to the less privileged in the community \cite{jooKeepOnlineOption2021,ninerBetterWhomLeveling2021}.
\rita{Sessions of an hybrid conference could have both chairs online, as attendee or speaker. This mean that is important when organizing the conference to include a third person who is in the venue to help the in-person participants, take the questions that are both online and in-person and communicate with the remote session chairs.}

In fact, experts in theorizing about and building communities of practice [Wenger, 1999] who have been holding hybrid events since 2011 [Wenger-Tayner, 2021] consider that inclusive hybrid events are the future. These authors emphasize the need that all participants of hybrid events must go through the same event quality to the point that in-person and online participants pay the same registration amount to participate in high-quality hybrid conferences. These authors recommend having a buddy system, where every person attending online has an in-person attendant that makes sure the online person can follow the event and participate in the event as if they were in the room [Wenger-Trayner, 2020]. Wenger-Trayner (2020, 2021) also recommends having a chat function for back-channel conversations between all attendants, a virtual space for all online attendants to talk to each other, and allocating time to small group conversations, often mixing in-person and online groups. \rita{The chat system of the conference can have a list of all the participants, with their registration name and if they are on-line or in-person, to facilitate the connections.}  The physical venue will have to include multiple ways to connect with online folks such as tablets on the meeting venue tables and cameras that allow online attendants to follow who is speaking in the in-person space. 
\rita{The venue selection must respond to some criteria such as give online support both bot in-person and online participants, to have stable and secure connection for the in-person participant to join the online event. Moreover, offer the option to create virtual room to connect in-person and online participant for networking events, such as scientific interest group, or LGBTQIA+-friendly spaces. Exhibition stands in the venue need to have facilities to connect with online participants (virtual stand). Furthermore, the poster session need to be carefully thought to give the best experience to both in-person and online participants. This should have both components to allow online participants to see the poster and have the possibility to ask questions. However, it also need allow who want to present their poster virtually to be seen by in-person participants and respond to questions about their works. Lastly to better plan your hybrid conference ask in advance who wish to travel to the venue and who wish to participate online. This will save you time and money for looking for a suitable venue.}
